Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2023-11-10 incidence-prevalence.tex
Various edits

I start by defining these terms in \cref{inc-prev:sec:definitions}.
I then move on to explaining the three processes that are involved: the \emph{infection process}, the \emph{prevalence process} and the \emph{observation process}.
The infection process (\cref{inc-prev:sec:infection-process}) models how infections arise.
The concepts here are well-established within infectious disease epidemiology (see\todo{ref intro chapter}).
The prevalence process (\cref{inc-prev:sec:prevalence-process}) describes the relationship between incidence and prevalence.
Here, I adapt previous work to the SARS-CoV-2 context.
In particular, the recoveries need explicit modelling which was not required for HIV/AIDS.
The observation process describes the relationship between prevalence and the observed data.
Unlike in HIV/AIDS backcalculation, the CIS has a sampling mechanism that needs to be modelled.
This model is guided by the specific study design of the CIS.
Therefore, the model is specific to this context, although much would be applicable to other, similarly designed prevalence surveys.
Finally, \cref{inc-prev:sec:MCMC} will introduce Markov chain Monte Carlo (MCMC), an inference technique I will make extensive use of in this thesis, before \cref{inc-prev:sec:conclusion} concludes the chapter.

@@

Time will be considered discrete in this chapter.
I use days, however, the results generalise in the obvious way to other time units.
I make this choice for pragmatic reasons.
Most epidemiological data is only available at daily granularity or, when more granular data exists, is often unreliable.
Furthermore, within-day patterns produce variations in transmission (\eg due to sleep) and data (\eg due to logistical considerations around collection); these variations are not of interest in most contexts.

@@

\emph{Prevalence} of a disease on day $t$, is the number of people with the disease on that day~\autocite[90]{lashModern}.
What it means to have a disease can vary by context (see\todo{ref section discussing what it means to have covid}).
I define an individual to have SARS-CoV-2 on day $t$ if, had they been swabbed on day $t$, that swab would have returned a positive PCR test.
I refer to these individuals as \emph{prevalent} individuals.
I assume that an individual's prevalent status cannot change over the course of a day.
Therefore, prevalence is constant over any single day $t$.
I denote the number of prevalent individuals on day $t$ as $P_t$.
Therefore, the prevalence is $P_t/\Npop$.

Duration, or time-to-event, distributions can be expressed in several equivalent ways.
The duration distribution I will be concerned with is the distribution of the number of days for which an individual is prevalent.
Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$.
I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$.
$D_i = E_i - B_i + 1$ is their \emph{duration of positivity}, the number of days for which they are prevalent.
I model each $D_i$ as a discrete independent and identically distributed (iid) random variable; extending the methodology to account for covariates is left to further work.
The duration is a positive integer, \ie $D_i \in \{1, 2, \dots\}$.
The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=0}^d f_{D}(i)$.
The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$~\autocite{yanDistribution}.
The \emph{hazard} at time $t$ is the probability of recovering at time $t$ conditional on having survived to time $t$: $\lambda(t) = \prob(D = t \mid D \geq t) = f_{D}(d) / S(d)$.
For discrete variables, the convention that the survival function is $\prob(D >d)$, rather than $\prob(D \geq d)$, is also commonly used, I adapt that latter convention.
Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be parameterised in terms of the hazard; these two quantities are related by $S(t) = \prod_{i=1}^{t-1} (1 - \lambda(i))$.

@@

A point process is a random function of continuous time $K(t)$.
$K(t)$ must take non-negative integer values and be non-decreasing in $t$.
The interpretation of $K(t)$ is the number of events, here infections, that have occurred in the interval $(-\infty, t]$~\autocites[244]{yanDistribution}.
In the context of a disease, the point process is discretised into incidence for integer $t$ as $Z_t = K(t) - K(t-1)$.

A counting process has \emph{independent increments} if the number of events in disjoint intervals are independent~\autocite[244]{yanDistribution}.
That is, for any $t_1 < t_2 < t_3 < t_4$, $K(t_2) - K(t_1)$ and $K(t_4) - K(t_3)$ are independent.

A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocites{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
A Poisson process is a point process with independent increments, and the property that $K(s) - K(t) \dist \Poi\left(\int^s_t \lambda(r) dr \right)$ where $\lambda(r) \geq 0$ is the \emph{intensity} of the epidemic at time $r$~\autocites[244]{yanDistribution}.
A time-homogenous Poisson process is the special case where $\lambda(r)$ is constant.
The independent increment property says that, conditional on $\lambda(r)$, the number of infections in disjoint intervals are independent.
This implies $Z_t$ and $Z_{t'}$ are independent for $t \neq t'$.
Extensions to this model can incorporate covariates, such as geography or age~\autocite[e.g.][]{diggleModeling}.

@@

Two features of a Poisson process are violated for epidemics.
Epidemics commonly exhibit overdispersion and have feedback loops~\autocite{beckerDependent}.
Overdispersion means that $\V(Z_t) > \E(Z_t)$ for any day $t$.
In a Poisson process these counts have a Poisson distribution, hence, $\V(Z_t) = \E(Z_t)$.
Feedback loops mean that the stochastically low or high numbers of infections, possibly due to overdispersion, affect the future number of infections.
This violates the assumption of independent increments.
In particular, $\lambda(t+\delta t)$ is dependent on $K(t)$ for $\delta t > 0$.
Both of these features violate the properties of a Poisson process.
Despite these issues, consistent inference on the incidence can still be produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.

A common cause of overdispersion is super-spreading events~\autocite{lloyd-smithSuperspreading}.
A super-spreading event is an event where many more people than average are infected.
That is, $Z_t >> \E(Z_t)$ for some $t$ during which a super-spreading event occurs.
Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocites{endoEstimating}{adamClustering}{mccloskeySARS}.
An infamous example of a SARS-CoV-2 super-spreading event was the Skagit County choir practice in March 2020~\autocite{hamnerHigh}.
One infected individual is thought to have infected at least 32 further individuals during a single choir practice.
For comparison, the average individual infected around three others in the early stages of the pandemic~\autocite{pellisChallenges}. 
Overdispersion can be incorporated by replacing the assumption that $K(s) - K(t)$ is Poisson distributed with assuming it is negative binomially distributed.
This would not affect $\E(K(s) - K(t))$, but the negative binomial introduces an additional overdispersion parameter, commonly notated $k$.
%The negative binomial tends to Poisson as $k\to\infty$.
Generally, $k<1$ is considered as significant levels of overdispersion.
Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, and may be decreased by some interventions including lockdowns~\autocites{quiltyReconstructing}{quiltyUnderstanding}.
If $K(s) - K(t)$ has a negative binomial distribution, then its standard deviation scales on the order of the mean.
If it is Poisson distributed, then the standard deviation is on the order of the square root of the mean.
A particular implication of this is that the coefficient of variation (the standard deviation divided by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.
The disadvantage of the negative binomial distribution is that it loses much of the mathematical convenience of a Poisson distribution.

The independence assumption of a Poisson or negative binomial process is violated because if the number of infections is stochastically high (or low), this feeds back to the process intensity at future times.
This can be seen simply in the event of a super-spreading event.
If a super-spreading event occurs, then many more individuals than suggested by the process intensity will be infected.
These additional infections will propagate the epidemic further, increasing the intensity at future times.
Point processes can be extended to \emph{branching processes} to account for this feedback~\autocite[246]{yanDistribution}, although these are beyond the scope of this thesis.

@@

Denote by $R_{t,t'}$ be the number of infection episodes with $B_i=t$ and $E_i=t'$.
For each infection beginning at $t$, the probability that it ends at $t'$ is $\prob(E_i = t' \mid B_i = t) = \prob(D_i = t' - t + 1 \mid B_i = t) = f_D(t' - t + 1)$.
Since these durations are assumed independent, the number of episodes of each duration is multinomially distributed~\autocite{paganoHIV}.
Therefore:
\begin{align}
\begin{bmatrix}
  R_{t,t} \\ R_{t,t+1} \\ \vdots \\ R_{t,t+d_\text{max}-1}
\end{bmatrix} \mid Z_t
\sim \MN \left(
  Z_t, 
  \begin{bmatrix}
    f_D(1) \\ f_D(2) \\ \vdots \\ f_D(d_\text{max})
  \end{bmatrix}
\right).
\end{align}
Now, denote by $P_{t_z,t_p}$ the number of infection episodes with $B_i = t_z$ and $E_i \geq t_p$.
Therefore, the individuals in which these infection episodes occur are prevalent at time $t_p$.
$P_{t_z,t_p}$ is $Z_{t_z}$ minus those that recover before $t_p$ (\ie at or up to $t_p - 1$).
Hence:
\begin{align}
    P_{t_z,t_p} = \begin{cases}
      0 &t_p < t_z\\
      Z_{t_z} - \sum_{i=0}^{t_p-t_z-1} R_{t_z,t_z+i} &t_z \leq t_p < t_z + d_\text{max}\\
      0 &t_p \geq t_z + d_\text{max}.
  \end{cases} \label{inc-prev:eq:Ptt-to-R}
\end{align}
Finally, the $P_t$ can be expressed as the total number of people prevalent at time $t$, regardless of when they were infected.
The limits of the sum are limited by the zeroes in \cref{inc-prev:eq:Ptt-to-R}.
\begin{align}
  P_t
  &= \sum_{t_z=-\infty}^\infty P_{t_z,t} \\
  &= \sum_{i=0}^{\dmax-1} P_{t-i,t} \label{inc-prev:eq:Pt-to-Ptt} \\
  &= \sum_{i=0}^{\dmax-1} \left(Z_{t_z} - \sum_{j=0}^{i-1} R_{t-i,t-i+j} \right) &\text{by \cref{inc-prev:eq:Ptt-to-R}}\label{inc-prev:eq:Pt-to-Rtt}.
\end{align}

@@

The Metropolis-Hastings algorithm~\autocite{hastingsMCMC} describes a general method for constructing such a Markov chain.
The algorithm requires a proposal distribution $q(\psi' \mid \psi)$, which proposes moving to a new state $\psi'$ from the current state $\psi$, and a calculation of $p(\psi \mid \vec{y})$ up to some constant, $\pi(\psi)$.
The proposal distribution is most commonly a Normal distribution centred on $\psi$, this is used in \cref{E-SEIR}.
The proposal is then either accepted or rejected based on the \emph{acceptance ratio}, which is computed using $\pi$.
The full algorithm is given in \cref{inc-prev:algorithm:MH}, with further details in available in a wide-range of sources~\autocites[e.g.][]{brooksMCMCNotes}[chapter 11]{gelmanBDA}.
\begin{algorithm}
 set $\psi^0$ to some initial value\;
 \For{$i = 1, \dots, M$}{
  sample a proposal $\psi' \sim q(\psi' \mid psi)$ \;
  calculate the acceptance probability $\alpha(\psi, \psi') = \min \left( 1, \frac{\pi(\psi') q(\psi \mid \psi')}{\pi(\psi) q(\psi' \mid \psi)} \right)$ \;
  with probability $\alpha(\psi, \psi')$ set $\psi^i$ to $\psi'$, otherwise set $\psi^i$ to $\psi^{i-1}$ \;
 }
 \caption{The Metropolis-Hastings algorithm.}
 \label{inc-prev:algorithm:MH}
\end{algorithm}

\todo[inline]{Talk about HMC and NUTS}

@@

Convergence is how quickly the Markov chain approaches the stationary distribution.
Early draws from the Markov chain reflect the choice of the starting point, $\psi^0$, rather than the stationary distribution that the chain will eventually converge to~\autocite[282]{gelmanBDA}.
The simplest technique for assessing convergence is graphically, using traceplots~\autocite[81]{brooksMCMCNotes}.
Traceplots show the value of $\psi^i$ against $i$.
If the chain has converged, then the traceplot will show a random walk around the stationary distribution.
This can be improved by using multiple chains with different starting points~\autocite[283]{gelmanBDA}.
If both chains have converged, then they will be in the same location.

@@

The stability of estimates considers the error arising because only a finite sample of the distribution of interest is available.
Furthermore, the sample will tend to be autocorrelated meaning that estimates using the example have higher variance than if it were an iid sample from the posterior~\autocite[286]{gelmanBDA}.
The \emph{effective sample size} corrects for the autocorrelation.
The effective sample size can be interpreted as the number of iid samples that would have the same variance as the MCMC sample~\autocite[286]{gelmanBDA}{vehtariRhat}.
Various techniques have been proposed for computing the effective sample size, I will follow the approach of \textcite{vehtariRhat}.

>>>>

I start by defining these terms in \cref{inc-prev:sec:definitions}.
I then move on to explaining the three processes that are involved: the \emph{infection process}, the \emph{prevalence process} and the \emph{observation process}.
The infection process (\cref{inc-prev:sec:infection-process}) models how infections arise.
The concepts here are well-established within infectious disease epidemiology (see\todo{ref intro chapter}).
The prevalence process (\cref{inc-prev:sec:prevalence-process}) describes the relationship between incidence and prevalence.
Here, I adapt previous work to the SARS-CoV-2 context.
In particular, recoveries need considering, which was not the case for HIV/AIDS.
The observation process describes the relationship between prevalence and the observed data.
Unlike in HIV/AIDS backcalculation, the CIS has a sampling mechanism that needs to be modelled.
This model is guided by the specific study design of the CIS.
Therefore, the model is specific to this context, although much would be applicable to other, similarly designed prevalence surveys.
Finally, \cref{inc-prev:sec:MCMC} will introduce Markov chain Monte Carlo (MCMC), an inference technique I will make extensive use of in this thesis.
\Cref{inc-prev:sec:conclusion} concludes the chapter.

@@

Time will be considered discrete in this chapter.
I use days, however, the results generalise in the obvious way to other time units.
I choose discrete time for pragmatic reasons.
Most epidemiological data is only available at daily granularity or, when more granular data exists, is often unreliable.
Furthermore, within-day patterns produce variations in transmission (\eg due to sleep) and data (\eg due to logistical considerations around collection); these variations are not of interest in most contexts.

@@

\emph{Prevalence} of a disease on day $t$, is the number of people with the disease on that day~\autocite[90]{lashModern}.
What it means to have a disease can vary by context (see\todo{ref section discussing what it means to have covid}).
Here, an individual is defined to have SARS-CoV-2 on day $t$ if, had they been swabbed on day $t$, that swab would return a positive PCR test.
I refer to these individuals as \emph{prevalent} individuals.
I assume that an individual's prevalent status cannot change over the course of a day.
Therefore, prevalence is constant over any single day $t$.
I denote the number of prevalent individuals on day $t$ as $P_t$.
Therefore, the prevalence is $P_t/\Npop$.

Duration, or time-to-event, distributions can be expressed in several equivalent ways.
The duration distribution I will be concerned with is the distribution of the number of days for which an individual is prevalent.
Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$.
I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$.
$D_i = E_i - B_i + 1$ is their \emph{duration of positivity}, the number of days for which they are prevalent.
I model each $D_i$ as a discrete independent and identically distributed (iid) random variable; extending the methodology to account for covariates is left to further work.
The duration is a positive integer, \ie $D_i \in \{1, 2, \dots \dmax \}$, where $\dmax$ is the maximum duration.
The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$ (in some of the literature on survival with discrete variables, the convention that the survival function is $\prob(D >d)$, rather than $\prob(D \geq d)$, is used).
The \emph{hazard} on day $d$ is the probability of recovering at time $d$ conditional on having survived to time $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be parameterised in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.

Since $D_i$ is a discrete random variable, $f_D(d) \geq 0$ for all $d$ and $\sum_{i=1}^{\dmax} f_D(i) = 1$.
This implies that $F_D(d)$ is monotonically increasing.
Therefore, $S(d)$ is monotonically decreasing.
Both $F_D$ are $S$ are in the interval $[0, 1]$.
$\lambda(d)$ is in the interval $[0, 1]$, and can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).

@@

A point process is a random function of continuous time $K(t)$.
$K(t)$ must take non-negative integer values and be monotonically increasing in $t$.
The interpretation of $K(t)$ is the number of events, here infections, that have occurred in the interval $(-\infty, t]$~\autocites[244]{yanDistribution}.
In the context of a disease, the point process can be discretised into the incidence on day $t$ as $Z_t = K(t) - K(t-1)$, where $t$ is an integer.

A counting process has \emph{independent increments} if the number of events in disjoint intervals are independent conditional on the intensity~\autocite[244]{yanDistribution}.
That is, for any $t_1 < t_2 < t_3 < t_4$, $K(t_2) - K(t_1)$ and $K(t_4) - K(t_3)$ are independent.

A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocites{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
A Poisson process is a point process with independent increments, and the property that $K(s) - K(t) \dist \Poi\left(\int^s_t \nu(r) dr \right)$ where $\nu(r) \geq 0$ is the \emph{intensity} of the epidemic at time $r$~\autocites[244]{yanDistribution}.
A time-homogenous Poisson process is the special case where $\nu(r)$ is constant.
The independent increment property says that, conditional on $\nu(r)$, the number of infections in disjoint intervals are independent.
This implies $Z_t$ and $Z_{t'}$ are independent for $t \neq t'$.
Extensions to this model can incorporate covariates, such as geography or age~\autocite[e.g.][]{diggleModeling}.

@@

Two features of a Poisson process are violated for epidemics.
Epidemics commonly exhibit \emph{overdispersion} and have feedback loops~\autocite{beckerDependent}.

Overdispersion means that $\V(Z_t) > \E(Z_t)$.
In a Poisson process $Z_t$ has a Poisson distribution, hence, $\V(Z_t) = \E(Z_t)$.
Therefore, overdispersed epidemics cannot be modelled by a Poisson process.

A common cause of overdispersion is super-spreading events~\autocite{lloyd-smithSuperspreading}.
A super-spreading event is an event where many more people than average are infected.
That is, $Z_t >> \E(Z_t)$ for some $t$ during which a super-spreading event occurs.
Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocites{endoEstimating}{adamClustering}{mccloskeySARS}.
An infamous example of a SARS-CoV-2 super-spreading event was the Skagit County choir practice in March 2020~\autocite{hamnerHigh}.
One infected individual is thought to have infected at least 32 further individuals during a single choir practice.
For comparison, the average individual infected around three others in the early stages of the pandemic~\autocite{pellisChallenges}. 
Overdispersion can be incorporated by replacing the assumption that $K(s) - K(t) \dist \Poi(\int_{t}^{s} \nu(r) dr)$ with $K(s) - K(t) \dist \NBc(\int_{t}^{s} \nu(r) dr, k)$ where $k$ is the overdispersion parameter, normally assumed to be constant absent interventions.
Note that the expected number of events has not changed, but the variance has increased (see \cref{E-distributions} for details of parameterisations and moments).
%The negative binomial tends to Poisson as $k\to\infty$.
Generally, $k<1$ is considered as significant levels of overdispersion.
Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, and may be decreased by some interventions including lockdowns~\autocites{quiltyReconstructing}{quiltyUnderstanding}.
If $K(s) - K(t)$ has a negative binomial distribution, then its standard deviation scales on the order of the mean.
If it is Poisson distributed, then the standard deviation is on the order of the square root of the mean.
A particular implication of this is that the coefficient of variation (the standard deviation divided by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.
The disadvantage of the negative binomial distribution is that it loses much of the mathematical convenience of a Poisson distribution.

Feedback loops mean that stochastically low or high numbers of infections affect the future number of infections.
That is, $\nu(t+\Delta t)$ is dependent on $K(t)$, even after conditioning on $\nu(t)$, for $\Delta t > 0$.
This violates the assumption of independent increments.
Both of these features violate the properties of a Poisson process.
Despite these issues, consistent inference on the incidence can still be produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.

The independence assumption of a Poisson or negative binomial process is violated by an infectious disease because stochastically high (or low) or low incidence, feeds back to the process intensity at future times.
This can be seen most obviously in the event of a super-spreading event.
If a super-spreading event occurs, then many more individuals than suggested by the process intensity will be infected.
These additional infections will propagate the epidemic further, increasing the intensity at future times.
Point processes can be extended to \emph{branching processes} to account for this feedback~\autocite[246]{yanDistribution}, although these are beyond the scope of this thesis.

@@

Denote by $R_{t,t'}$ the number of infection episodes with $B_i=t$ and $E_i=t'$.
For each infection beginning at $t$, the probability that it ends at $t'$ is $\prob(E_i = t' \mid B_i = t) = \prob(D_i = t' - t + 1 \mid B_i = t) = f_D(t' - t + 1)$.
Since these durations are assumed independent, the number of episodes of each duration is multinomially distributed~\autocite{paganoHIV}.
Therefore:
\begin{align}
\begin{bmatrix}
  R_{t,t} \\ R_{t,t+1} \\ \vdots \\ R_{t,t+d_\text{max}-1}
\end{bmatrix} \mid Z_t
\sim \MN \left(
  Z_t, 
  \begin{bmatrix}
    f_D(1) \\ f_D(2) \\ \vdots \\ f_D(d_\text{max})
  \end{bmatrix}
\right).
\end{align}
Now, denote by $P_{t_z,t_p}$ the number of individuals infected at $t_z$ and prevalent at $t_p$.
Formally, $P_{t_z,t_p}$ is the infection episodes with $B_i = t_z$ and $E_i \geq t_p$.
$P_{t_z,t_p}$ is $Z_{t_z}$ minus those that recover before $t_p$ (\ie at or up to $t_p - 1$).
Hence:
\begin{align}
    P_{t_z,t_p} = \begin{cases}
      0 &t_p < t_z\\
      Z_{t_z} - \sum_{i=0}^{t_p-t_z-1} R_{t_z,t_z+i} &t_z \leq t_p < t_z + d_\text{max}\\
      0 &t_p \geq t_z + d_\text{max}.
  \end{cases} \label{inc-prev:eq:Ptt-to-R}
\end{align}
Finally, the $P_t$ can be expressed as the total number of people prevalent at time $t$, regardless of when they were infected.
The limits of the sum are limited by the zeroes in \cref{inc-prev:eq:Ptt-to-R}.
\begin{align}
  P_t
  &= \sum_{t_z=-\infty}^\infty P_{t_z,t} \\
  &= \sum_{i=0}^{\dmax-1} P_{t-i,t} \label{inc-prev:eq:Pt-to-Ptt} \\
  &= \sum_{i=0}^{\dmax-1} \left(Z_{t_z} - \sum_{j=0}^{i-1} R_{t-i,t-i+j} \right) &\text{by \cref{inc-prev:eq:Ptt-to-R}}\label{inc-prev:eq:Pt-to-Rtt}.
\end{align}

@@

The Metropolis-Hastings algorithm~\autocite{hastingsMCMC} describes a general method for constructing such a Markov chain.
The algorithm requires a proposal distribution $q(\psi' \mid \psi)$, which proposes moving to a new state $\psi'$ from the current state $\psi$, and a calculation of $p(\psi \mid \vec{y})$ up to some constant, $\pi(\psi)$.
The proposal distribution is most commonly a (possibly multivariate) Normal distribution centred on $\psi$, this is used in \cref{E-SEIR}.
The proposal is then either accepted or rejected based on the \emph{acceptance ratio}, which is computed using $\pi$.
Tuning the proposal distribution to propose ``good'' moves is a key part of MCMC.
The ideal proposal distribution is the posterior of interest, however, this is impossible because that is what MCMC is trying to estimate~\autocite[296]{gelmanBDA}.
One alternative is \emph{adaptive} algorithms which tune the proposal distribution based on previous draws in an attempt to approximate the posterior; I use such a scheme in \cref{E-SEIR:sec:MCMC}.
The full Metropolis-Hastings algorithm is given in \cref{inc-prev:algorithm:MH}, with further details in available in a wide-range of sources~\autocites[e.g.][]{brooksMCMCNotes}[chapter 11]{gelmanBDA}.
\begin{algorithm}
 set $\psi^0$ to some initial value\;
 \For{$i = 1, \dots, M$}{
  sample a proposal $\psi' \sim q(\psi' \mid psi)$ \;
  calculate the acceptance probability $\alpha(\psi, \psi') = \min \left( 1, \frac{\pi(\psi') q(\psi \mid \psi')}{\pi(\psi) q(\psi' \mid \psi)} \right)$ \;
  with probability $\alpha(\psi, \psi')$ set $\psi^i$ to $\psi'$, otherwise set $\psi^i$ to $\psi^{i-1}$ \;
 }
 \caption{The Metropolis-Hastings algorithm.}
 \label{inc-prev:algorithm:MH}
\end{algorithm}

MCMC using a Normal proposal distribution can be inefficient.
In particular, it often fails to propose moves into the tails of the posterior distribution\todo{any suggested refs for this (Paul?) - we've discussed this lots but I've never read it anywhere}.
\emph{Hamiltonian Monte Carlo} (HMC)~\autocites{nealMCMC}{nealImproved}{duaneHybrid} improves the proposals by using information from the gradient of the posterior density (with respect to the parameters).
HMC simulates a physical system travelling through the posterior space with momentum.
The momentum increases in the direction the posterior density increases, using the gradient.
The \emph{No U-Turn Sampler} (NUTS)~\autocite{hoffmanNUTS} is a variant of HMC that automatically tunes the number of steps to take in the Hamiltonian dynamics.
NUTS has been implemented in a variety of software packages making it easy to apply to arbitrary posterior distributions.
I will make use of the implementation in the package \emph{Sampling through adaptive neighborhoods} (Stan)~\autocite{Stan-2-32-2}.
Stan implements \emph{automatic differentiation} to calculate the gradient of the posterior density without any user input.
Automatic differentiation uses the chain rule to calculate the gradient of a function from its code by breaking it down into a series of elementary operations, for which the gradient is known.
For a more detailed introduction to HMC, NUTS, and Stan see \textcite[chapter 12]{gelmanBDA}.

@@

Convergence is how quickly the Markov chain approaches the stationary distribution.
Early draws from the Markov chain reflect the choice of the starting point, $\psi^0$, rather than the stationary distribution that the chain will eventually converge to~\autocite[282]{gelmanBDA}.
The simplest technique for assessing convergence is graphically, using traceplots~\autocite[81]{brooksMCMCNotes}.
Traceplots show the value of $\psi^i$ against $i$.
If the chain has converged, then the traceplot will show a random walk around the stationary distribution.
This can be improved by using multiple chains with different starting points~\autocite[283]{gelmanBDA}.
If all chains have converged, then they will be in the same location.

@@

The stability of estimates considers the error arising because only a finite sample of the distribution of interest is available.
Furthermore, the sample will tend to be autocorrelated meaning that estimates using the example have higher variance than if it were an iid sample from the posterior~\autocite[286]{gelmanBDA}.
The \emph{effective sample size} corrects for the autocorrelation.
The effective sample size can be interpreted as the number of iid samples that would have the same variance as the MCMC sample~\autocites[286]{gelmanBDA}{vehtariRhat}.
Various techniques have been proposed for computing the effective sample size, I will follow the approach of \textcite{vehtariRhat}, implemented in RStan~\autocite{RStan-2-32-3}.
